# ECCB workflow {#eccb_workflow}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

TODO:

- rename the workflow
- 

THINGS TO CONSIDER
- We should probably use same dataset
- 

PRELIMINARY SCHEDULE

13:00-13:20 Leo
 - Idea: Introduction and TreeSE
 - Introduction to Bioconductor
 - Overview of data containers
 - Introduction to TreeSE
13:20-13:25 short Q&A
 - Participants have time to explore on their own
13:25-13:30 short break
 - Time to set up for the next session
 
13:30-13:50 Pande
 - Idea: After introducing TreeSE, we can now use methods from mia.
 - What are diversity and composition?
 - How to calculate and plot alpha diversity indices?
 - How to calculate and plot PCoA(?)
13:50-13:55 short Q&A
 - Participants have time to explore on their own
13:55-14:00 short break
 - Time to set up for the next session
 
14:00-14:20 Tuomas
 - Idea: Now participants know that we can utilize methods from mia. Introduce slots and show how to handle complex data.
 - reducedDim slot was briefly introduced in Pande's session. (That highlights the idea of slots hopefully since the result is the same as already calculated.)
 - Introduce agglomerateByRanks() and altExp slot
 - Add another omics to altExp and demonstrating column subsetting to show sample linkage
 - Calculate cross-association?
14:20-14:30 short Q&A
 - Participants have time to explore on their own

15:00-15:20 Stefanie
 - Idea: Show what additional capabilities there are
 - Maybe: we can calculate simple correlation network based on cross-association results (Tuomas's session) but this can be further extended?
15:20-15:25 short Q&A
 - Participants have time to explore on their own
15:25-15:30 short break
 - Time to set up for the next session
 
15:30-15:50 Giulio
 - Idea: Show what additional capabilities there are
 - Maybe: the analyses performed earlier can be done in graphical interface?
15:50-15:55 short Q&A
 - Participants have time to explore on their own
15:55-16:00 short break
 - Time to set up for the next session



## Introduction

Explain the workflow: what user can find from here?

## Load packages

```{r}
#| label: install
#| output: FALSE
# List of packages that we need
packages <- c("mia",  "miaViz", "scater")

# Get packages that are already installed
packages_already_installed <- packages[ packages %in% installed.packages() ]

# Get packages that need to be installed
packages_need_to_install <- setdiff( packages, packages_already_installed )

# Loads BiocManager into the session. Install it if it is not already installed.
if( !require("BiocManager") ){
    install.packages("BiocManager")
    library("BiocManager")
}

# If there are packages that need to be installed, installs them with BiocManager
# Updates old packages.
if( length(packages_need_to_install) > 0 ) {
   install(packages_need_to_install, ask = FALSE)
}

# Load all packages into session. Stop if there are packages that were not
# successfully loaded
pkgs_not_loaded <- !sapply(packages, require, character.only = TRUE)
pkgs_not_loaded <- names(pkgs_not_loaded)[ pkgs_not_loaded ]
if( length(pkgs_not_loaded) > 0 ){
    stop(
        "Error in loading the following packages into the session: '",
        paste0(pkgs_not_loaded, collapse = "', '"), "'")
}
```

## Importing data

Introduce the dataset briefly.

```{r}
#| label: loadDataset
data("Tito2024QMP")
tse <- Tito2024QMP
```

## Bioconductor resources for microbiome data science (Leo)

TreeSE

- Introduction to TreeSE
- Transformations?

## Analysis of community diversity and composition (Pande)
 - Idea: After introducing TreeSE, we can now use methods from mia.
 - What are diversity and composition?
 - How to calculate and plot alpha diversity indices?
 - How to calculate and plot PCoA(?)
 
##  Microbiome data integration (Tuomas)

- Idea: Now participants know that we can utilize methods from mia. Introduce slots and show how to handle complex data.
 - reducedDim slot was briefly introduced in Pande's session. (That highlights the idea of slots hopefully since the result is the same as already calculated.)
 - Introduce agglomerateByRanks() and altExp slot
 - Add another omics to altExp and demonstrating column subsetting to show sample linkage
 - Calculate cross-association?
 
 ## Microbial network analysis (Stefanie)
 
 ## Interactive microbiome data exploration with iSEEtree (Giulio)
