# ECCB workflow {#eccb_workflow}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

TODO:

- rename the workflow
- 

THINGS TO CONSIDER
- We should probably use same dataset. Data integration part requires multiomics data. 
- I added idea for story but feel free to modify.

PRELIMINARY SCHEDULE

13:00-13:20 Leo
 - Idea: Introduction and TreeSE
 - Introduction to Bioconductor
 - Overview of data containers
 - Introduction to TreeSE
13:20-13:25 short Q&A
 - Participants have time to explore on their own
13:25-13:30 short break
 - Time to set up for the next session
 
13:30-13:50 Pande
 - Idea: After introducing TreeSE, we can now use methods from mia.
 - What are diversity and composition?
 - How to calculate and plot alpha diversity indices?
 - How to calculate and plot PCoA(?)
13:50-13:55 short Q&A
 - Participants have time to explore on their own
13:55-14:00 short break
 - Time to set up for the next session
 
14:00-14:20 Tuomas
 - Idea: Now participants know that we can utilize methods from mia. Introduce slots and show how to handle complex data.
 - reducedDim slot was briefly introduced in Pande's session. (That highlights the idea of slots hopefully since the result is the same as already calculated.)
 - Introduce agglomerateByRanks() and altExp slot
 - Add another omics to altExp and demonstrating column subsetting to show sample linkage
 - Calculate cross-association?
14:20-14:30 short Q&A
 - Participants have time to explore on their own

15:00-15:20 Stefanie
 - Idea: Show what additional capabilities there are
 - Maybe: we can calculate simple correlation network based on cross-association results (Tuomas's session) but this can be further extended?
15:20-15:25 short Q&A
 - Participants have time to explore on their own
15:25-15:30 short break
 - Time to set up for the next session
 
15:30-15:50 Giulio
 - Idea: Show what additional capabilities there are
 - Maybe: the analyses performed earlier can be done in graphical interface?
15:50-15:55 short Q&A
 - Participants have time to explore on their own
15:55-16:00 short break
 - Time to set up for the next session



## Introduction

Explain the workflow: what user can find from here?

## Load packages

```{r}
#| label: install
#| output: FALSE
# List of packages that we need
packages <- c("ComplexHeatmap", "mia",  "miaViz", "scater", "shadowtext")

# Get packages that are already installed
packages_already_installed <- packages[ packages %in% installed.packages() ]

# Get packages that need to be installed
packages_need_to_install <- setdiff( packages, packages_already_installed )

# Loads BiocManager into the session. Install it if it is not already installed.
if( !require("BiocManager") ){
    install.packages("BiocManager")
    library("BiocManager")
}

# If there are packages that need to be installed, installs them with BiocManager
# Updates old packages.
if( length(packages_need_to_install) > 0 ) {
   install(packages_need_to_install, ask = FALSE)
}

# Load all packages into session. Stop if there are packages that were not
# successfully loaded
pkgs_not_loaded <- !sapply(packages, require, character.only = TRUE)
pkgs_not_loaded <- names(pkgs_not_loaded)[ pkgs_not_loaded ]
if( length(pkgs_not_loaded) > 0 ){
    stop(
        "Error in loading the following packages into the session: '",
        paste0(pkgs_not_loaded, collapse = "', '"), "'")
}
```

## Importing data

There are several openly available datasets listed in [@sec-example-data]. In
this workflow, we will retrieve data from the curatedMetagenomicData resource.
We'll focus on a dataset that includes both colorectal cancer (CRC) patients
and control subjects. Our goal is to investigate whether CRC is associated
with alterations in the gut microbiota.

```{r}
#| label: import_dataset

# Study from which we want to fetch data
study <- "GuptaA_2019"
# Experiments that we want to fetch
experiments <- c("pathway_abundance", "relative_abundance") |>
    paste0(collapse = "|")
search_pattern <- paste0(study, ".(", experiments, ")")

# Fetch data
data_list <- curatedMetagenomicData(
    search_pattern, dryrun = FALSE, counts = TRUE, rownames = "short")

# The assay name for taxonomy profile is incorrect since we have counts, not
# relative abundances
exp_ind <- grep("relative_abundance", names(data_list))[[1]]
tse <- data_list[[exp_ind]]
assayNames(tse) <- "counts"
data_list[[exp_ind]] <- tse

# For pathway data, extract pathway information from rownames and add them to
# rowData
exp_ind <- grep("pathway", names(data_list))[[1]]
tse <- data_list[[exp_ind]]
rowData(tse)[["full_name"]] <- rownames(tse)
rowData(tse)[["pathway"]] <- sub("\\|.*", "", rowData(tse)[["full_name"]])
data_list[[exp_ind]] <- tse

# Get taxonomy object
exp_ind <- grep("relative_abundance", names(data_list))[[1]]
tse <- data_list[[exp_ind]]
tse
```

```{r}
#| label: agglomeration

tse_family <- agglomerateByRank(tse, rank = "family")
```

```{r}
#| label: transformation

tse <- transformAssay(tse, method = "relabundance")
tse <- transformAssay(tse, method = "clr", pseudocount = 1)

tse_family <- transformAssay(tse_family, method = "relabundance")
tse_family <- transformAssay(tse_family, method = "clr", pseudocount = 1)
```


## Bioconductor resources for microbiome data science (Leo)

Idea for the story:

"
TreeSE is the primary data container used in miaverse framework.It is extension to SE.
It works like this. mia package has methods for common operations. For instance, 
we can tranform the data.
"


## Analysis of community diversity and composition (Pande)

Idea for the story:

"
After preprocessing the data, we can continue to analysis. It is common to analyze the alpha diversity
of samples. ...

To analyze similarity between samples, we can perform PCoA. ...
"

##  Microbiome data integration (Tuomas)

Linking to Pande's session

"

As we show, CRC patients seems to have altered microbiota. However, that is only association and does not tell about causality or mechanisms. To study complex mechanisms, we need more data. --> More data, we need efficient ways to handle data from multiple experiments. --> Show altExp can be utilized to handle multiple experiments

We show cross-association analysis; very simple "data integrtation" method. More in OMA.

"

```{r}
altExp(tse, "family") <- tse_family
```


```{r}
altExp(tse, "pathway") <- data_list[[1]]

altExp(tse, "pathway") <- agglomerateByVariable(altExp(tse, "pathway"), by = "rows", "pathway")
altExp(tse, "pathway") <- agglomerateByPrevalence(altExp(tse, "pathway"), assay.type = "pathway_abundance", prevalence = 0.2, detection = 0.001)
altExp(tse, "pathway") <- transformAssay(altExp(tse, "pathway"), assay.type = "pathway_abundance", method = "clr", pseudocount = TRUE)


res <- getCrossAssociation(
    tse,
    altexp1 = "family", "pathway",
    assay.type1 = "clr", assay.type2 = "clr", mode = "matrix", test.signif = TRUE)


# Function for marking significant correlations with "X"
add_signif <- function(j, i, x, y, width, height, fill) {
    # If the p-value is under threshold
    if( !is.na(res$p_adj[i, j]) & res$p_adj[i, j] < 0.05 ){
        # Print "X"
        grid.shadowtext(
            sprintf("%s", "X"), x, y, gp = gpar(fontsize = 8, col = "#f5f5f5"))
    }
}

# Create a heatmap
p <- Heatmap(
    res$cor,
    # Print values to cells
    cell_fun = add_signif,
    heatmap_legend_param = list(
     title = "correlation", legend_height = unit(5, "cm")),
    column_names_rot = 45
)
p
```

"Now we have calculated cross-association between microbes and predicted metabolic pathways. As seen, certain bacteria seems to be responsible for certain metabolic acitivity."


## Microbial network analysis (Stefanie)
 
Idea for the story:

"
We performed cross-association analysis in previous session. Analyzing microbial
networks, however, usually requires more sophisticated apporoaches. 

"

## Interactive microbiome data exploration with iSEEtree (Giulio)

Idea for the story:

"
For interactive analysis, iSEEtree is available. It allows to perform common analyses in graphical interface.

"
