# ECCB workflow {#eccb_workflow}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

TODO:

- rename the workflow
- 

THINGS TO CONSIDER
- We should probably use same dataset. Data integration part requires multiomics data. 
- I added idea for story but feel free to modify.

PRELIMINARY SCHEDULE

13:00-13:20 Leo
 - Idea: Introduction and TreeSE
 - Introduction to Bioconductor
 - Overview of data containers
 - Introduction to TreeSE
13:20-13:25 short Q&A
 - Participants have time to explore on their own
13:25-13:30 short break
 - Time to set up for the next session
 
13:30-13:50 Pande
 - Idea: After introducing TreeSE, we can now use methods from mia.
 - What are diversity and composition?
 - How to calculate and plot alpha diversity indices?
 - How to calculate and plot PCoA(?)
13:50-13:55 short Q&A
 - Participants have time to explore on their own
13:55-14:00 short break
 - Time to set up for the next session
 
14:00-14:20 Tuomas
 - Idea: Now participants know that we can utilize methods from mia. Introduce slots and show how to handle complex data.
 - reducedDim slot was briefly introduced in Pande's session. (That highlights the idea of slots hopefully since the result is the same as already calculated.)
 - Introduce agglomerateByRanks() and altExp slot
 - Add another omics to altExp and demonstrating column subsetting to show sample linkage
 - Calculate cross-association?
14:20-14:30 short Q&A
 - Participants have time to explore on their own

15:00-15:20 Stefanie
 - Idea: Show what additional capabilities there are
 - Maybe: we can calculate simple correlation network based on cross-association results (Tuomas's session) but this can be further extended?
15:20-15:25 short Q&A
 - Participants have time to explore on their own
15:25-15:30 short break
 - Time to set up for the next session
 
15:30-15:50 Giulio
 - Idea: Show what additional capabilities there are
 - Maybe: the analyses performed earlier can be done in graphical interface?
15:50-15:55 short Q&A
 - Participants have time to explore on their own
15:55-16:00 short break
 - Time to set up for the next session



## Introduction

Explain the workflow: what user can find from here?

## Load packages

```{r}
#| label: install
#| output: FALSE
# List of packages that we need
packages <- c("mia",  "miaViz", "scater")

# Get packages that are already installed
packages_already_installed <- packages[ packages %in% installed.packages() ]

# Get packages that need to be installed
packages_need_to_install <- setdiff( packages, packages_already_installed )

# Loads BiocManager into the session. Install it if it is not already installed.
if( !require("BiocManager") ){
    install.packages("BiocManager")
    library("BiocManager")
}

# If there are packages that need to be installed, installs them with BiocManager
# Updates old packages.
if( length(packages_need_to_install) > 0 ) {
   install(packages_need_to_install, ask = FALSE)
}

# Load all packages into session. Stop if there are packages that were not
# successfully loaded
pkgs_not_loaded <- !sapply(packages, require, character.only = TRUE)
pkgs_not_loaded <- names(pkgs_not_loaded)[ pkgs_not_loaded ]
if( length(pkgs_not_loaded) > 0 ){
    stop(
        "Error in loading the following packages into the session: '",
        paste0(pkgs_not_loaded, collapse = "', '"), "'")
}
```

## Importing data

Introduce the dataset briefly.

```{r}
#| label: loadDataset
#| echo: false
data("HintikkaXOData")

# Get MAE object and store biomarker experiment to altExp of microbiota
mae <- HintikkaXOData
tse <- getWithColData(mae, 1)
altExp(tse, "biomarkers") <- getWithColData(mae, 3)
```

## Bioconductor resources for microbiome data science (Leo)

Idea for the story:

"
TreeSE is the primary data container used in miaverse framework.It is extension to SE.
It works like this. mia package has methods for common operations. For instance, 
we can tranform the data.
"


## Analysis of community diversity and composition (Pande)

Idea for the story:

"
After preprocessing the data, we can continue to analysis. It is common to analyze the alpha diversity
of samples. ...

To analyze similarity between samples, we can perform PCoA. ...
"

##  Microbiome data integration (Tuomas)

Idea for the story:

"
Microbiome data is complex, which requires efficient way to handle and process
the data. With large datasets, not only is it crucial to manage the data
effectively, but it's also important to store and access intermediate and final
results efficiently as it was done in previous alpha and beta diversity calculations.

TreeSE offers a robust framework for organizing microbiome data, accommodating
various layers of information, including taxonomic ranks and sample metadata.
This structured approach facilitates easier data management and analysis.

To incorporate additional omics data, the altExp slot within TreeSE can be
utilized. This slot maintains the linkage between samples across different data
types. For even more complex datasets, the MultiAssayExperiment framework can
be employed, providing advanced functionality for integrating and analyzing diverse omics datasets.

OMA also shows how to integrate different omics sets in analysis,
cross-association analysis being the most simple approach.

"

## Microbial network analysis (Stefanie)
 
Idea for the story:

"
We performed cross-association analysis in previous session. Analyzing microbial
networks, however, usually requires more sophisticated apporoaches. 

"

## Interactive microbiome data exploration with iSEEtree (Giulio)

Idea for the story:

"
For interactive analysis, iSEEtree is available. It allows to perform common analyses in graphical interface.

"
